---
title: "orchrd: Workflow Orchestration for R"
subtitle: "A complete guide for beginner, intermediate, and advanced users"
author: "orchrd package"
description: >
  A practical guide to workflow orchestration, automation,
  filesystem watching, system execution, Docker operations,
  and deployment with orchrd.
---

```{r setup, include=FALSE}  
knitr::opts_chunk$set(  
  echo = TRUE,  
  eval = FALSE,          # examples are illustrative; do not run at build time  
  collapse = TRUE,  
  comment = "#>"  
)  
```  
  
# Introduction {#introduction}  
  
`orchrd` brings PowerShell's workflow philosophy into R: simple commands,  
clean structured outputs, pipeline thinking, retry logic, and  
automation-friendly logging. It is designed as infrastructure for the  
openafrR ecosystem but is useful for any R developer orchestrating  
scripts, APIs, and system processes.  
  
This guide is organized in three tiers so you can enter at your level:  
  
- **Beginner** -- run a command, chain a few steps, validate data.  
- **Intermediate** -- logging, retries, config, secrets, notifications.  
- **Advanced** -- background jobs, batching, rollbacks, checkpoints,  
  Docker scaffolding and diagnostics.  
  
## Installation {#installation}  
  
```{r}  
# install.packages("remotes")  
remotes::install_github("laws2020/orchrd")  
  
library(orchrd)  
```  
  
## The two core return types {#return-types}  
  
Almost everything in `orchrd` returns one of two predictable shapes, so  
you never have to guess what you got back.  
  
| Class              | Returned by                     | Key fields                                  |  
|--------------------|---------------------------------|---------------------------------------------|  
| `orchrd_result`    | `run_cmd()`, `run_script()`     | `$ok`, `$status`, `$stdout`, `$stderr`      |  
| `orchrd_pipeline`  | `pipe_exec()`                   | `$ok`, `$steps` (data frame), `$elapsed_sec`|  
  
Both have `print()` methods, so inspecting them in the console is clean.  
  
---  
  
# Beginner {#beginner}  
  
## Running a single command: `run_cmd()` {#run-cmd}  
  
`run_cmd()` spawns an external process via `processx::run()` and returns a  
structured `orchrd_result`. You never parse raw text or guess exit codes.  
  
```{r}  
res <- run_cmd("git", c("status", "--short"))  
  
res$ok        # TRUE if exit status was 0  
res$status    # integer exit code  
res$stdout    # character vector of stdout lines  
res$stderr    # character vector of stderr lines  
```  
  
**Why not `system2()`?** `run_cmd()` gives you separated stdout/stderr, a  
timeout, and a consistent object you can test with `res$ok` instead of  
inspecting attributes.  
  
## Running a script file: `run_script()` {#run-script}  
  
`run_script()` detects the file extension and invokes the right  
interpreter automatically -- R, Python, PowerShell, bash, or Quarto.  
  
```{r}  
run_script("analysis.R")          # -> Rscript --vanilla  
run_script("clean.py")            # -> python3  
run_script("deploy.ps1")          # -> pwsh -NoProfile -NonInteractive -File  
run_script("report.qmd")          # -> quarto render  
```  
  
You never need to remember the exact interpreter flags; one call handles  
all of them and returns the same `orchrd_result` as `run_cmd()`.  
  
## Chaining steps: `pipe_exec()` {#pipe-exec}  
  
`pipe_exec()` is the heart of the package. Each step is a one-sided  
formula (`~ expr`) or a named formula (`name = ~ expr`). The special  
symbol `.x` holds the output of the previous step.  
  
```{r}  
result <- pipe_exec(  
  fetch   = ~ read.csv("input.csv"),  
  clean   = ~ subset(.x, !is.na(value)),  
  summary = ~ aggregate(value ~ group, .x, mean)  
)  
  
result$ok            # TRUE if every step succeeded  
result$steps         # data frame: name, ok, elapsed, message  
result$elapsed_sec   # total run time  
```  
  
Each step is timed, logged, and error-handled. Inspect the run with  
`print(result)`.  
  
## Validating data inline: `validate_data()` and asserts {#validate}  
  
Quick, inline assertions that fail loudly with clear messages. Not a  
replacement for a full framework -- meant to catch obvious problems early.  
  
```{r}  
validate_data(  
  df,  
  schema = ~ assert_schema(.x, c("id", "value", "group")),  
  bounds = ~ assert_range(.x, "value", min = 0, max = 100),  
  filled = ~ assert_no_na(.x, c("id", "value"))  
)  
```  
  
- `assert_schema(data, cols)` -- required columns must exist.  
- `assert_range(data, col, min, max)` -- numeric column within bounds.  
- `assert_no_na(data, cols)` -- no missing values in the named columns.  
  
Wrap each rule as a formula (`~ assert_*()`) so `validate_data()` can  
collect **all** failures instead of stopping at the first.  
  
---  
  
# Intermediate {#intermediate}  
  
## Observability: `log_step()`, `with_log()`, `log_config()`, `get_log()` {#logging}  
  
Make any workflow observable. Log at a severity level, print via `cli`,  
and optionally write NDJSON to a file for later analysis.  
  
```{r}  
log_config(level = "debug", file = "run.ndjson")   # configure the session logger  
  
with_log("nightly refresh", {  
  log_step("Starting fetch", level = "info")  
  data <- fetch_data()  
  log_step("Fetched rows", level = "debug", data = list(n = nrow(data)))  
})  
  
get_log()   # data frame of timestamp, level, message for the session  
```  
  
`get_log()` returns a tidy data frame; the full structured `.data` per  
entry is preserved in the NDJSON sink for `jsonlite::stream_in()`.  
  
## Retrying flaky work: `retry()` and `retry_labeled()` {#retry}  
  
Wrap any expression with automatic retry logic: exponential backoff,  
random jitter, a `max_wait` cap, and pattern-matched error conditions.  
  
```{r}  
retry(  
  httr2::req_perform(req),  
  times    = 5,  
  wait     = 2,       # 2s, 4s, 8s ... (doubles)  
  max_wait = 30,      # cap the delay  
  jitter   = TRUE,    # +/- 30% so parallel callers do not sync up  
  on       = "timeout|HTTP 5"   # only retry matching errors  
)  
  
# Labeled variant prints a header, ideal inside pipe_exec():  
pipe_exec(  
  a = ~ retry_labeled("Source A", get_data("a")),  
  b = ~ retry_labeled("Source B", get_data("b"))  
)  
```  
  
## Configuration: the `config_*` family {#config}  
  
Read and manage layered configuration for a pipeline.  
  
```{r}  
config_get("api_url")     # fetch one key  
config_has("token")       # TRUE / FALSE  
config_all()              # everything as a list  
config_clear()            # reset  
```  
  
## Secrets: `secret()`, `reveal()`, `is_secret()` {#secrets}  
  
Read secrets from environment variables with three guarantees plain  
`Sys.getenv()` does not give you: presence validation, log safety, and  
masked printing.  
  
```{r}  
key <- secret("AZURE_KEY")   # errors immediately if unset  
key                          # prints as <secret: AZURE_KEY>, never the value  
is_secret(key)               # TRUE  
  
reveal(key)                  # the only way to get the plain string  
```  
  
Because printing and `as.character()` are masked, secrets will not leak  
through `str()`, `dput()`, logs, or accidental console echo. Pass  
`reveal(key)` explicitly to any function that needs the raw value.  
  
## Completion signals: `notify()` {#notify}  
  
Signal when an unattended pipeline finishes. Supports desktop  
notifications, Slack, Microsoft Teams, and plain console output.  
  
```{r}  
pipe_exec(  
  build  = ~ run_script("build.R"),  
  deploy = ~ run_script("deploy.R")  
) |>  
  notify(channel = "slack", webhook = reveal(secret("SLACK_WEBHOOK")))  
```  
  
Non-console channels fall back to the console on failure, so a broken  
webhook never silently swallows the completion signal.  
  
## Pre-flight checks: `need()` and `health_check()` {#preflight}  
  
Fail fast, before a single step runs.  
  
```{r}  
# Assert the environment is ready (commands, env vars, files, packages)  
need(  
  commands = c("git", "docker"),  
  env      = c("AWS_REGION"),  
  files    = c("config.yml"),  
  packages = c("httr2", "jsonlite")  
)  
  
# Probe that external services are reachable  
health_check(c(  
  "https://api.example.com",  
  "db.internal:5432",  
  "docker"  
))  
```  
  
`need()` lists **every** missing item at once (not just the first) and  
includes install hints for well-known commands. `health_check()` returns  
a tidy data frame with a `type` column and `response_ms` timings.  
  
---  
  
# Advanced {#advanced}  

## Reactive automation {#reactive} 
  
`orchrd` also watches the world and reacts to it: filesystem events and  
cron-style time events, both from a single in-session API. Watchers are  
backed by a compiled C++ layer (inotify on Linux, ReadDirectoryChangesW  
on Windows, kqueue on macOS); schedules run on R's event loop via `later`.  
  
## React to files: `fire_on()` and the `on_*()` verbs {#watchers}  
  
`fire_on(event, path, action, ...)` binds an action to a filesystem  
event. `action` is either a function `(event, path)` or a one-sided  
formula `~ expr` where `.path` and `.event` are bound.  
  
```{r}  
# Formula shorthand  
on_arrive("~/nbs_drops/", pattern = "\\.xlsx$",  
          action = ~ message("File arrived: ", .path))  
  
# Function form  
on_change("~/configs/pipeline.yml",  
          action = function(event, path) reload_config(path))  
```  
  
Event verbs: `on_arrive()` (created), `on_change()` (modified),  
`on_leave()` (deleted), `on_rename()` (renamed). Each returns a watcher  
ID invisibly. Low-level `watch_dir()` / `watch_file()` sit underneath.

## React to time: `on_schedule()` {#schedule}  
  
Register a recurring in-session timer with a 5-field cron expression  
(`min hour dom mon dow`). It never blocks the session and needs no  
system cron. Requires the `later` package.  
  
```{r}  
on_schedule("0 8 * * *",  action = ~ fetch_cbn_rates())        # 08:00 daily  
on_schedule("*/15 9-17 * * 1-5", action = ~ log_step("beat"))  # every 15m, Mon-Fri  
```

## Safety conditions: `guard()`, `disk_ok()`, `mem_ok()`, `net_ok()` {#guards}  
  
`guard(id, condition, message, on_fail)` wraps a watcher's callback so it  
only fires when `condition()` is `TRUE`; otherwise the event is logged  
and `on_fail` (`"warn"`, `"stop"`, or `"silent"`) is taken.  
  
```{r}  
id <- on_arrive("~/cbn_feeds/", action = ~ fetch_rates(.path))  
  
guard(id, condition = ~ disk_ok(threshold = 0.90),  
      message = "Disk > 90% full - skipping fetch")  
guard(id, condition = ~ net_ok(), message = "No network")  
```

## Rate control: `throttle()` {#throttle}  
  
Debounce fires at the trailing edge of a burst; `throttle(id, ms)` fires  
immediately on the first event then suppresses re-fires for `ms`.  
  
```{r}  
throttle(id, ms = 30000)   # fire once, ignore re-drops for 30s  
```  
  
## Manage watchers {#manage-watchers}  
  
```{r}  
watcher_list()      # active watchers as a data frame  
watcher_status(id)  # one watcher's state  
watcher_pause(id); watcher_resume(id)  
watcher_log()       # recent events (FIRED / GUARDED / THROTTLED / errors)  
unwatch(id)         # stop one  
unwatch_all()       # stop all  
```  
  
## Pipeline bridge: `pipe_on_arrive()`, `pipe_on_change()`, `health_report()` {#reactive-bridge}  
  
Dispatch a full `orchrd` pipeline in the background the moment a file  
lands, and print a system-health summary.  
  
```{r}  
pipe_on_arrive("~/drops/", pipeline = function(.file) run_script(.file))  
health_report()     # disk / memory / network status  
```


## Background jobs: `run_bg()` and `await()` {#background}  
  
`run_bg()` launches a job in a background R process (via `callr`) and  
returns immediately; `await()` blocks until it finishes and collects the  
result. These are the primitives `batch_exec()` and tripwire's  
`orchrd::run_bg()` bridge build on.  
  
```{r}  
job <- run_bg(~ heavy_computation(big_data))  
# ... do other work ...  
result <- await(job)          # blocks until done, returns the value  
```  
  
## Controlled parallelism: `batch_exec()` and `batch_combine()` {#batch}  
  
Run a list of tasks N at a time until exhausted -- for example, fetching  
54 country datasets without overwhelming the network.  
  
```{r}  
tasks <- lapply(country_codes, function(code) {  
  function() fetch_country(code)  
})  
  
results <- batch_exec(tasks, workers = 4)   # 4 at a time  
combined <- batch_combine(results)          # rbind the per-task frames  
```  
  
`batch_exec()` spawns via `run_bg()` and cleans up leftover jobs (kill or  
`await`) if it stops early.  
  
## Rollback on failure: `with_rollback()` and `deploy_step()` {#rollback}  
  
For deployment-style workflows: if a step fails, undo what it did before  
the error propagates. If the step succeeds, the rollback never runs.  
  
```{r}  
pipe_exec(  
  upload = ~ deploy_step(  
    deploy_expr = upload_to_server("out.parquet"),  
    revert_expr = delete_from_server("out.parquet"),  
    label       = "upload parquet"  
  )  
)  
```  
  
## Resumable pipelines: checkpoints {#checkpoints}  
  
`checkpoint()`, `step_done()`, `resume_pipeline()`, and  
`clear_checkpoint()` let a long pipeline skip already-completed steps on  
a re-run.  
  
```{r}  
checkpoint("fetch")           # mark progress  
step_done("fetch")            # query whether a step already ran  
resume_pipeline()             # continue from the last checkpoint  
clear_checkpoint()            # start fresh  
```  
  
## Artifacts: `store_artifact()`, `delete_artifact()`, `push_aws()` {#artifacts}  
  
Persist and move intermediate outputs.  
  
```{r}  
store_artifact(model, "model-v1")  
push_aws("model-v1", bucket = "s3://my-bucket/models")  
delete_artifact("model-v1")  
```  
  
## Metrics: `pipeline_summary()` {#metrics}  
  
Aggregate performance across multiple runs by reading the NDJSON logs  
written by `log_config(file=)` and/or the JSON run logs from  
`pipe_exec(.log_file=)`.  
  
```{r}  
summ <- pipeline_summary("logs/")  
summ$runs           # number of runs  
summ$success_rate   # fraction of steps that succeeded  
summ$elapsed        # timing distribution  
summ$errors         # data frame of warn/error entries  
```  
  
## Docker scaffolding {#docker-scaffold}  
  
Generate reproducible Docker assets for any project stack.  
  
```{r}  
generate_dockerfile(stack = "r-shiny")   # stack-aware Dockerfile  
optimize_build_speed()                   # project .Rprofile with CRAN mirror  
minify_r_dependencies()                  # leaner Rprofile.site  
secure_secrets()                         # .dockerignore excluding secrets  
```  
  
## Docker diagnostics and maintenance {#docker-ops}  
  
Cross-platform Docker operations (the Docker CLI is invoked directly, not  
through PowerShell) plus Windows-only recovery helpers.  
  
```{r}  
diagnose()                          # overall Docker health report  
check_docker_oom()                  # containers killed for out-of-memory  
clean_docker_storage()              # docker system prune (careful: volumes)  
force_remove_container("trsbs-api") # docker rm -f  
free_docker_port(8080)              # Windows: find and free a held port  
debug_portal_api("portal", "api")  # inspect networking + recent logs  
  
# Windows-only Docker Desktop recovery:  
repair_docker()                     # restart Docker Desktop + WSL  
fix_docker_permissions()            # add user to docker-users group  
```  
  
The `repair_docker()` and `fix_docker_permissions()` helpers guard on OS  
and error clearly on non-Windows systems.  
  
---  
